---
title: "Lesson 1: R Basics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```



## Basic arithmetic, assignment 

The normal mathematic operators work in R, and they follow the order of operations:

```{r}
1 / 20 * 30
```

You can use parentheses to change the order in which the expressions are calculated:

```{r}
1 / (20 * 30)
```

```{r}
cos(2 * pi)
```

You can assign values to objects with `<-`, the assignment operator. The shortcut for this is:

* For Windows Users, `Alt` + `-`
* For Mac Users, `Cmd` + `-`

After assignment, we can perform whatever operations we would like with that variable:

```{r}
x <- 5
x + 5
```

```{r}
sin(2 * x)
```

```{r}
x * 500 / sin(x)
```

R also has the modulus operator `%%`, which returns the remainder when `x` is divided by `y`. For example:

```{r}
5 %% 2
```


### Exercise 1.1

* How can you tell when a number is divisible by 3?
* How can you tell when a number is divisible by 10?
* How can you tell when a number is odd?

## Using functions

Every programming language has functions. In R, they look something like:

```{r, eval = FALSE}
function_name(arg_1 = ..., arg_2 =..., ...)
```

For example, the `seq` function takse a starting number, an ending number, and an interval, and returns a sequence of numbers based on those inputs. For example, to count to 100 by 5:

```{r}
seq(0, 100, 5)
```

And just like before, we can assign this to a variable and do other stuff with it. For example, we can add a scalar to each component:

```{r}
y <- seq(0, 100, 5)
y + 2
```

We can multiply the sequence by a scalar:

```{r}
y * 4
```

Another function is the `length` function. We can use it to check the number of objects contained in an object:

```{r}
length(y)
```

We can do component-wise addition of two scalars:

```{r}
z <- seq(1, 20)
y + z
```

### Exercise 1.2

* What was the warning we received when we added `y` and `z` together? How did R handle it?
* When we made the list of numbers `y`, we provided a third number to indicate the interval to count by. When we made `z`, we left that out.

## Accessing elements 

We can access elements of the sequence we created earlier by using bracket notation. For example, to access the second element of `y`, we can do:

```{r}
y[2]
```

To access the first 5 elements, we can do:

```{r}
y[1:5]
```


## Working with data 

Most of the time you're working with R, you'll be manipulating some data that was given to you, like expression data, an OTU abundance table, or yield data. The most common form of data is some sort of rectangular container, such as a data frame, a table, a tibble, a matrix, etc. While there are some differences between all of them, they all function similarly. Note that we're going to begin our investigation using Base R functionality before we move on to using `tidyverse` (specifically, `dplyr`) functions.

We're going to begin learning how to manipulate data by using `mtcars`, a dataset that comes packaged with R.

Let's learn more about the `mtcars` dataset:

```{r, eval = FALSE}
?mtcars
```

So there's information about a few different cars here. Let's take a look at the data:

```{r}
mtcars
```

Most times, datasets are too large to be printed out to the whole screen at once and we just want to get a look at some of the data to get a feel for it. We can use `head` or `tail` to look at the first or last few rows of a dataset, respectively:

```{r}
head(mtcars)
```

```{r}
tail(mtcars)
```

We can check how many rows and columns there are by doing nrows and nols:

```{r}
nrow(mtcars)
```

```{r}
ncol(mtcars)
```

Alternatively, we can use `dim`, which returns the number of rows and columns as a pair of numbers:

```{r}
dim(mtcars)
```

We can use notation similar to what we used earlier to access particular elements of the data frame. For instance, if I wanted the value in the 5th column of the first row, I could access it like this:

```{r}
mtcars[1, 5]
```

To access the first fives values in the first row, you can do this:

```{r}
mtcars[1, 1:5]
```

And to get the whole row, you can leave the second argument empty:

```{r}
mtcars[1, ]
```

Alternatively, you can access columns by their name by putting the name in quotes:

```{r}
mtcars[1, 'cyl']
```

To access multiple column names in this way, you need to *combine* (or *concatenate*) them together with the `c` function:

```{r}
mtcars[1, c('mpg', 'cyl', 'disp')]
```

You can do the same kind of syntax to access the rows of the data frame:

```{r}
mtcars[1:5, 1:5]
```

```{r}
mtcars["Toyota Corolla", 1:5]
```

You can also use negative selection to deselect columns and rows. For instance, if we didn't care about the `qsec` column, we can do:

### Exercise 1.3

* Use `rownames` to find out all the different cars that are in the dataset. Print out the rows of these cars.

```{r}
mtcars[c("Toyota Corolla", "Honda Civic", "Datsun 710"), ]
```


* Print out the number of cylinders, quarter-mile time, number of carburetors, and whether or not the engine is v-shaped for the cars you chose above. See the documentation for `?mtcars` to figure out which columns have that information. 

```{r}
mtcars[c("Toyota Corolla", "Honda Civic", "Datsun 710"), c("cyl", "qsec", "carb", "vs")]
```


## Subsetting data based on conditions

Let's say we're interested in cars with 4 cylinders. The following statement returns a vector of `TRUE` and `FALSE` indicating the rows in which the statement `cylinders equals 4` is true.

```{r}
mtcars$cyl == 4
```

We can then use this vector to subset `mtcars` so that it shows only the cars with 4 cylinders:

```{r}
mtcars[mtcars$cyl == 4, ]
```

Similarly, we can use this to look at cars that have more than 4 cylinders:

```{r}
mtcars[mtcars$cyl > 4, ]
```


We can use `|`, the boolean `or` operator, to look at cars that are 4 or 6 cylinders:

```{r}
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
```

And we can combine it with our indexing from earlier to look at the number of cylinders and quarter-mile speed for each of these cars:

```{r}
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, c("cyl", "qsec")]
```

### Exercise 1.4

* Use the `&`, the boolean `and` operator, to subset `mtcars` by those cars with 6 cylinders and less than 35 miles per gallon.
* Use the `arrange` function 


## Getting started with the tidyverse

Everything we did up to now was done using base R and was useful in helping you get a grasp of how data frames work. However, we're now going to start using the tidyverse suite of functions from here out. If you do some searching online, you'll find some people with strong opinions about whether or not to use the Tidyverse functions. There are arguments on both sides. For example, On the one hand, Tidyverse functions make certain things a lot easier, like quickly prototyping and editing data analysis pipelines, and it is often easier to read Tidyverse code than base R code. On the other hand, code written using Tidyverse functions are not 100% portable, because your collaborator may not have installed the packages required. We'll be using the Tidyverse functions when we can, but we will use Base R functionality as required.

Let's revisit what we did earlier in the subsetting section. We'll start by importing the tidyverse functions:

```{r}
library(tidyverse)
```

The first function we will learn about is `select`. We can use the `select` function to select variables in a dataframe. For example, to select the number of cylinders and miles per gallon, we can do:

```{r}
select(mtcars, 'cyl', 'mpg')
```

We can also use `select` to *deselect* certain variables; when used in this way, `select` will return all the columns except for the variables indicated. For example, if we aren't interested in miles per gallon, number of cylinders, or displacement, we can do:

```{r}
select(mtcars, -"mpg", -"cyl", -"disp")
```

You can even do something like this:

```{r}
select(mtcars, contains("p"))
```

You can still use column numbers in the `select` function:

```{r}
select(mtcars, 1:3)
```


The next function we're going to learn about is the `filter` function. This function subsets the data by certain conditions. To use the `filter` function to select cars with 4 cylinders, we would do:

```{r}
filter(mtcars, cyl == 4)
```

We can combine conditions just like before:

```{r}
filter(mtcars, (cyl == 4 | cyl == 8) & mpg < 25)
```

We can also combine the functions. For example, to see the miles per gallon, displacement force, and number of cylinders for all the cars with four cylinders, we can do:

```{r}
filter(select(mtcars, 'mpg', 'cyl', 'disp'), cyl == 4)
```

Note that `filter` discards rownames for `data.frames`. This is because the row names are not themselves a column. We'll work later with `tibbles`, which are functionally similar to `data.frames`, but have a few enhancements to make working with them easier. For now, in cases where the rownames are important, you can use the `subset` function in place of filter:

```{r}
subset(select(mtcars, 'mpg', 'cyl', 'disp'), cyl == 4)
```

`filter` and `subset` are mostly the same and we'll interchange them as needed in this tutorial. If you'd like to read more about the differences between the two, you can check out this [Stack Overflow post](https://stackoverflow.com/a/39882777/6024276).

Note that the code above got complicated very quickly. You can imagine that as you do more and more functions on the same line, it gets easier to misplace a parentheses or a quote and it becomes harder to read. To remedy this, you could use intermediate steps, where you save the result of each function call into a new object:

```{r}
mtcars.mpg_cyl_disp <- select(mtcars, 'mpg', 'cyl', 'disp')
mtcars.mpg_cyl_disp.four_cylinders <- subset(mtcars.mpg_cyl_disp, cyl == 4)
```

However, this quickly clutters up your workspace with objects and functions that you don't necessarily need.

The solution to this is to start using `%>%`, the pipe operator. This operator takes the object on the left and passes it as the input to the function call on the right. Liberal use of `%>%` and proper formatting will make your code much easier to read. For example, the above code can be rewritten with the pipe as:

```{r}
mtcars %>% 
  select('mpg', 'cyl') %>% 
  subset(cyl == 4)
```

We can immediately see that this is far more readable and easier to edit. For example, we can quickly add an additional `subset` call to get only those cars with higher than 25 miles-per-gallon:

```{r}
mtcars %>% 
  select('mpg', 'cyl') %>% 
  subset(cyl == 4) %>% 
  subset(mpg > 25)
```

It's a bit of a pain to type `%>%` by hand over and over, so get comfortable with the shortcut command:

* For windows users: `control` + `Shift` + `m`
* For Mac users: `Cmd` + `Shift` + `m`

You can find a full list of shortcuts under `Tools` > `Keyboard Shortcuts Help` or [here](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts).

You can still use everything you learned earlier 

```{r}
mtcars[c("Toyota Corolla", "Honda Civic", "Datsun 710"), ] %>% 
  select("mpg", "cyl")
```



### Exercise 1.5

For this exercise, we're going to use the data in the `nycflights13` package. Let's take a look at the data:

```{r}
library(nycflights13)
head(flights)
```

Take a few moments to read through the documentation with `?flights` and do the following tasks:

* How many flights were there on January 1, 2013?
* How many flights were operated by United or Delta?
* How many flights arrived more than two hours late?
* How many flights from JFK departed more than two hours late?
* Read about the `arrange` function. Use it to find the flights that arrived latest.

## Adding new variables, summarizing data:

We're usually interested in summarizing a data set. For example, we might be interested in the means of each of the variables in `mtcars`. One way to do this is to use the mean function on each of the columns individually:

```{r}
mean(mtcars$mpg)
# mean(mtcars$cyl)
# mean(mtcars$disp)
# etc...
```

Alternatively, we can just use `colMeans` on the entire data frame at once:

```{r}
colMeans(mtcars)
```

Another useful function is `summary`:

```{r}
summary(mtcars)
```


From here, we can then look at which cars have higher than average mpg:

```{r}
mtcars %>% 
  subset(mpg > mean(mtcars$mpg)) 
```

And we can sort them by `mpg`:

```{r}
mtcars %>% 
  subset(mpg > mean(mtcars$mpg)) %>% 
  arrange(mpg) %>% 
  head()
```

Again, because we're working with a `data.frame`, the dplyr functions drop the the row names. To preserve the row names, we can first reorder the `data.frame` using Base R, then subset it:

```{r}
mtcars[rev(order(mtcars$mpg)), ] %>% 
  subset(mpg > mean(mtcars$mpg)) %>% 
  head()
```

Let's suppose that we wanted to get the average `mpg` of cars by the number of cylinders it has. One way we can do this is to filter by the number of cylinders, select the `mpg` column, and then take the mean:

```{r}
mtcars %>% 
  filter(cyl == 4) %>% 
  select(mpg) %>% 
  summarise(mean_mpg = mean(mpg))
```

Note `summarise`, the new function that we used. `summarise` calculates whatever formula you provide on the data that is passed to it. We'll see shortly how can we use this to do some more interesting things.

That's great that we got the mean mpg for the cars with four cylinders, but we still want to know the mean mpg of the other cars. We could  repeat this for each of the number of cylinders to get what we need. However, this is not scalable and relies on knowing all the different levels of the variable you're looking at. This is not feasible for larger datasets. To remedy this, we can use the `group_by` and `summarise` commands from `tidyverse`. You pass the variable you would like group the original data by to the `group_by` function, and from there, we can use the summarise function to get some summary statistics by passing different formulas to the function call. 

For instance, to get the mean mpg by cylinder count, you can do:

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean = mean(mpg))
```

Here, we separated the data into groups based on the number of cylinders in the car, then performed the same calculation as above. 

We can also add a count column:

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean = mean(mpg),
            count = n())
```

Note that if we were just interested in the number of observations per group, we can simply use `count`:

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  count()
```


### Exercise 1.6

For the `mtcars` dataset:

* What is the average horsepower, weight, and displacement by cylinder count?
* How many cars are there with automatic transmission? 
* What is the count of cars by gear count?
* What is the breakdown of car count by transmission type **and** gear count? Hint: you can group by more than one variable at a time. Notice anything interesting?

For the `flights` dataset:

* Which carrier had the most flights that departed later than two hours?
* which carrier had the lowest average arrival delay?
* which destination airport had the most arrivals later than 30 minutes?
* Use `n_distinct` to find out how many distinct carriers and destinations there are. 
* Which route (origin-destination pair) had the most flights?
* Which day had the most flights?
* Which day had the arrival delays greater than 30 minutes?

## Adding new variables with mutate

The `mutate` function lets us add new variables to a data object by specifying a formula. For instance, to add a new column to `mtcars` showing the weight of the car in pounds (instead of thousands of pounds), we can do:

```{r}
mtcars %>% 
  mutate(weight_in_pounds = 1000 * wt) %>% 
  head()
```

We can add multiple variables at once. For example:

```{r}
mtcars %>% 
  mutate(weight_in_pounds = 1000 * wt,
         hp_per_cylinder = hp / cyl) %>% 
  head()
```

And from there, we can do all the other things that we have learned so far:

```{r}
mtcars %>% 
  mutate(weight_in_pounds = 1000 * wt,
         hp_per_cylinder = hp / cyl) %>% 
  select(hp, cyl, hp_per_cylinder) %>% 
  arrange(desc(hp_per_cylinder))
```



### Exercise 1.7

For the `mtcars`:

* What is the average `hp_per_cylinder` by cylinder count?

For the `flights` dataset:

* Create a new variable `gain`, defined as the the difference between the departure delay and arrival delay of a flight. What does this variable represent? What is the average gain for all flights in 2013? By airline?
* Which routes had the worst (ie, largest) average gain?


Next section: [ggplot2 basics](https://pommevilla.github.io/p3.bootcamp.r.2019/lesson_2.html)

